### 2022-10-24 dissertation/chapters/06-bayesian-hyper-heuristic.tex
🚧 (chapter 3): Just before DE

Section \ref{sec:bhh:selection_mechanism:random_events} presented the reader with the random events that are observed by the \ac{BHH} and Section \ref{sec:bhh:selection_mechanism:independence} provided an argument for independence between events. Chapter \ref{chap:probability} presented Bayes Theorem, but for convenience, it is given again in Equation \ref{eq:bhh:selection_mechanism:bayes_theorem} below.

@@

As was indicated in Section \ref{sec:bhh:selection_mechanism:independence}, Equation \ref{eq:bhh:selection_mechanism:bayes_theorem} can be evaluated for proportionality since the \ac{BHH} is not concerned with actual accurate selection probabilities, but rather implements a Bayesian classifier. The resulting proportionality is expressed in Equation \ref{eq:bhh:selection_mechanism:bayes_theorem_prop_to} below

@@

Notice how the Equation \ref{eq:bhh:selection_mechanism:bayes_theorem_prop_to} implements a Bayesian conditionally. This conditionality can be extended to include multiple criteria. The next section shows how the Bayesian classifier is implemented by providing the detail around the implemented predictive model.

@@

The parameterised predictive model, as derived from Bayes theorem presented in Equation \ref{eq:bhh:selection_mechanism:bayes_theorem} is then given in Equation \ref{eq:bhh:selection_mechanism:predictive_model} below.

@@

Notice the joint probability over events $E$ and $C$ given the selection of a heuristic $H$ denoted by the sums of the products of the separate parts, for $E$ and $C$, in the denominator. The calculation in the denominator can be intractable as has been previously indicated. However, in order to derive the posterior distribution as given in Equation \ref{eq:bhh:selection_mechanism:bayes_theorem}), one only has to evaluate to proportionality as is given in Equation \ref{eq:bhh:selection_mechanism:predictive_model_prop_to} below.

@@

The independence between events for the class label $H$ simply yields the \ac{PMF} of the Multinomial distribution as presented in Equation \ref{eq:bhh:selection_mechanism:naive_bayes:h_pmf} below.

@@

The independence between events $E$ given class label $H$ is denoted by the likelihood of $E$
conditional to the occurrence of heuristic $k$ and model parameter $\phi$ as
presented in Equation \ref{eq:bhh:selection_mechanism:naive_bayes:EgH_pmf} below.

@@

Finally, the independence between events for the performance criteria feature $C$ given class
label $H$ is denoted by the likelihood of $C$ conditional to the occurrence of
heuristic $k$ and model parameter $\psi$ as given in Equation \ref{eq:bhh:selection_mechanism:naive_bayes:CgH_pmf} below.

@@

Where $N_{k}$ is the summary variable as described for Equation \ref{eq:bhh:selection_mechanism:naive_bayes:CgH_pmf} above.
$N_{1,k}$ is a summary variable such that $N_{1,k} = \sum_{i=1}^{I}
      \mathbbm{1}_{1}(c_{i,k})\mathbbm{1}_{1}(h_{i,k})$, denoting the count of
occurrences of the events $c_{i}$ taking on a success $(c_{i}=1)$ and $h_{i}$
taking on class $k$, i.e. the count of occurrences of both succeeding in the
performance criteria and heuristic $k$ occurring together in $I$ independent,
identical runs. Similarly $N_{0,k} = N_{k} - N_{1,k}$ would denote the count of
occurrences of the events $c_{i}$ taking on a failure $(c_{i}=0)$ and $h_{i}$
taking on class $k$.

Given the assumptions of the naïve Bayes classifier, one could supplement Equations \ref{eq:bhh:selection_mechanism:naive_bayes:h_pmf}, \ref{eq:bhh:selection_mechanism:naive_bayes:EgH_pmf} and \ref{eq:bhh:selection_mechanism:naive_bayes:CgH_pmf} above into the proportional evaluation of the predictive model as given in Equation \ref{eq:bhh:selection_mechanism:predictive_model_prop_to}. This is presented in Equation \ref{eq:bhh:selection_mechanism:naive_bayes:HgEC_pmf} below.

@@

Computationally the equation presented in Equation \ref{eq:bhh:selection_mechanism:naive_bayes:HgEC_pmf} will underflow if the resulting probabilities are very small.

@@

When evaluating for Equation \ref{eq:bhh:selection_mechanism:naive_bayes:HgEC_pmf}, the numerical stability is shown to underflow if the resulting probabilities from its parts are very small. Multiplying multiple fractional parameters leads to an even smaller fractional number. Probabilities might be very low at some points during training. Consider an example where training has stagnated, effectively leading to a scenario where a credit assignment strategy never fulfills a credit assignment, yielding an extremely small probability for $\psi$. A solution to this problem is to apply the \textit{log-sum-exp trick}, parameterising distributions no longer using logits rather than probabilities. The transformation of Equation \ref{eq:bhh:selection_mechanism:naive_bayes:HgEC_pmf} using the log-sum-exp trick is given below in Equation \ref{eq:bhh:selection_mechanism:numerical_stability:log_sum_exp}.

@@

The calculations of the \ac{MLE} for $\hat{\theta}_{k}$, $\hat{\phi}_{j,k}$ and
$\hat{\psi}_{k}$ are now presented. The log likelihood of $\hat{\theta}$, $\hat{\phi}$ and $\hat{\psi}$ as derived
from the Equation \ref{eq:bhh:selection_mechanism:predictive_model_prop_to} is given in Equation \ref{eq:bhh:optimisation_step:mle:log_likelihood_all} below.

@@

Equation \ref{eq:bhh:optimisation_step:mle:log_likelihood_all} can be broken down into each of its factors. Consider the log likelihood of $\psi$  as denoted by Equation \ref{eq:bhh:optimisation_step:mle:log_likelihood_psi} below.

@@

The MLE for $\hat{\psi_{k}} $ can be calculated by taking the derivative of Equation \ref{eq:bhh:optimisation_step:mle:log_likelihood_psi} with respect to $\psi_{k}$ and equating to zero as presented in Equation \ref{eq:bhh:optimisation_step:mle:proof_log_likelihood_psi} as follows.

@@

The value for $\hat{\theta_{k}} $ can be calculated similarly. However, one has to compensate for the $K-1$ simplex $S$ by adding error factor $\epsilon$ to correct values for  $\theta$ where $\sum_{k}^{K} \theta_{k} \neq 1$. The new log likelihood function is then given in Equation \ref{eq:bhh:optimisation_step:mle:proof_log_likelihood_theta_part_1} below.

@@

Solving first for $\epsilon$ yields Equation \ref{eq:bhh:optimisation_step:mle:proof_log_likelihood_theta_part_2} below.

@@

Then solving for $\theta_{k}$ yields Equation \ref{eq:bhh:optimisation_step:mle:proof_log_likelihood_theta_part_3} as presented below.

@@

Substituting back into the Equation yields Equation \ref{eq:bhh:optimisation_step:mle:proof_log_likelihood_theta_part_4} below.

@@

The value for $\hat{\phi}_{j, k} $ is calculated very similarly. To compensate for the $K-1$ simplex $S$ an error factor $\lambda = (\lambda_{1}, \dots, \lambda_{J})$ is added. This follows in Equation \ref{eq:bhh:optimisation_step:mle:proof_log_likelihood_phi_part_1} as is presented below.

@@

Solving first for $\lambda_{j}$ yields Equation \ref{eq:bhh:optimisation_step:mle:proof_log_likelihood_phi_part_2} as is given below.

@@

Then solving for $\phi_{j,k}$ yields Equation \ref{eq:bhh:optimisation_step:mle:proof_log_likelihood_phi_part_3} below.

@@

The likelihood is presented in Equation \ref{eq:bhh:optimisation_step:map:bayesian_analysis_lamens} above. The event $H$ is a multinomial distribution with parameter $\theta$. It is known that the conjugate prior to a multinomial distribution is a Dirichlet distribution (see Chapter \ref{chap:probability}. The \textit{prior} probability distribution for  $\theta_{k}$ is thus presented in Equation \ref{eq:bhh:optimisation_step:map:map_theta_prior} below.

@@

Furthermore, the event $E$ is also a multinomial distribution with parameter $\phi$. The \textit{prior} probability distribution for $\phi_{j,k}$ is thus presented in Equation \ref{eq:bhh:optimisation_step:map:map_phi_prior} below.

@@

Finally, the event $C$ is a binomial distribution with parameter $\psi$. It is known that the conjugate prior to a binomial distribution is a Beta distribution (see Chapter \ref{chap:probability}). The prior probability distribution for $\psi_{k}$ is thus presented in Equation \ref{eq:bhh:optimisation_step:map:map_psi_prior} below.

@@

Putting the likelihood and priors together yields the posterior distribution as given in Equation \ref{eq:bhh:optimisation_step:map:map_psi_prior} below.

@@

From Equation \ref{eq:bhh:optimisation_step:map:posterior} above, it can be seen that the posterior distribution has the same form $\mathcal{A}(v)$ as the prior distribution. The concentration update operations can then be given in equations \ref{eq:bhh:optimisation_step:map:alpha_update_operation}, \ref{eq:bhh:optimisation_step:map:beta_update_operation}, \ref{eq:bhh:optimisation_step:map:gamma1_update_operation} and \ref{eq:bhh:optimisation_step:map:gamma2_update_operation} below.

>>>>

Section \ref{sec:bhh:selection_mechanism:random_events} presented the reader with the random events that are observed by the \ac{BHH} and Section \ref{sec:bhh:selection_mechanism:independence} provided an argument for independence between events. Chapter \ref{chap:probability} presented Bayes Theorem, but for convenience, it is given again in Equation~\eqref{eq:bhh:selection_mechanism:bayes_theorem} below.

@@

As was indicated in Section \ref{sec:bhh:selection_mechanism:independence}, Equation~\eqref{eq:bhh:selection_mechanism:bayes_theorem} can be evaluated for proportionality since the \ac{BHH} is not concerned with actual accurate selection probabilities, but rather implements a Bayesian classifier. The resulting proportionality is expressed in Equation~\eqref{eq:bhh:selection_mechanism:bayes_theorem_prop_to} below

@@

Notice how the Equation~\eqref{eq:bhh:selection_mechanism:bayes_theorem_prop_to} implements a Bayesian conditionally. This conditionality can be extended to include multiple criteria. The next section shows how the Bayesian classifier is implemented by providing the detail around the implemented predictive model.

@@

The parameterised predictive model, as derived from Bayes theorem presented in Equation~\eqref{eq:bhh:selection_mechanism:bayes_theorem} is then given in Equation~\eqref{eq:bhh:selection_mechanism:predictive_model} below.

@@

Notice the joint probability over events $E$ and $C$ given the selection of a heuristic $H$ denoted by the sums of the products of the separate parts, for $E$ and $C$, in the denominator. The calculation in the denominator can be intractable as has been previously indicated. However, in order to derive the posterior distribution as given in Equation~\eqref{eq:bhh:selection_mechanism:bayes_theorem}), one only has to evaluate to proportionality as is given in Equation~\eqref{eq:bhh:selection_mechanism:predictive_model_prop_to} below.

@@

The independence between events for the class label $H$ simply yields the \ac{PMF} of the Multinomial distribution as presented in Equation~\eqref{eq:bhh:selection_mechanism:naive_bayes:h_pmf} below.

@@

The independence between events $E$ given class label $H$ is denoted by the likelihood of $E$
conditional to the occurrence of heuristic $k$ and model parameter $\phi$ as
presented in Equation~\eqref{eq:bhh:selection_mechanism:naive_bayes:EgH_pmf} below.

@@

Finally, the independence between events for the performance criteria feature $C$ given class
label $H$ is denoted by the likelihood of $C$ conditional to the occurrence of
heuristic $k$ and model parameter $\psi$ as given in Equation~\eqref{eq:bhh:selection_mechanism:naive_bayes:CgH_pmf} below.

@@

Where $N_{k}$ is the summary variable as described for Equation~\eqref{eq:bhh:selection_mechanism:naive_bayes:CgH_pmf} above.
$N_{1,k}$ is a summary variable such that $N_{1,k} = \sum_{i=1}^{I}
      \mathbbm{1}_{1}(c_{i,k})\mathbbm{1}_{1}(h_{i,k})$, denoting the count of
occurrences of the events $c_{i}$ taking on a success $(c_{i}=1)$ and $h_{i}$
taking on class $k$, i.e. the count of occurrences of both succeeding in the
performance criteria and heuristic $k$ occurring together in $I$ independent,
identical runs. Similarly $N_{0,k} = N_{k} - N_{1,k}$ would denote the count of
occurrences of the events $c_{i}$ taking on a failure $(c_{i}=0)$ and $h_{i}$
taking on class $k$.

Given the assumptions of the naïve Bayes classifier, one could supplement Equations \ref{eq:bhh:selection_mechanism:naive_bayes:h_pmf}, \ref{eq:bhh:selection_mechanism:naive_bayes:EgH_pmf} and \ref{eq:bhh:selection_mechanism:naive_bayes:CgH_pmf} above into the proportional evaluation of the predictive model as given in Equation~\eqref{eq:bhh:selection_mechanism:predictive_model_prop_to}. This is presented in Equation~\eqref{eq:bhh:selection_mechanism:naive_bayes:HgEC_pmf} below.

@@

Computationally the equation presented in Equation~\eqref{eq:bhh:selection_mechanism:naive_bayes:HgEC_pmf} will underflow if the resulting probabilities are very small.

@@

When evaluating for Equation~\eqref{eq:bhh:selection_mechanism:naive_bayes:HgEC_pmf}, the numerical stability is shown to underflow if the resulting probabilities from its parts are very small. Multiplying multiple fractional parameters leads to an even smaller fractional number. Probabilities might be very low at some points during training. Consider an example where training has stagnated, effectively leading to a scenario where a credit assignment strategy never fulfills a credit assignment, yielding an extremely small probability for $\psi$. A solution to this problem is to apply the \textit{log-sum-exp trick}, parameterising distributions no longer using logits rather than probabilities. The transformation of Equation~\eqref{eq:bhh:selection_mechanism:naive_bayes:HgEC_pmf} using the log-sum-exp trick is given below in Equation~\eqref{eq:bhh:selection_mechanism:numerical_stability:log_sum_exp}.

@@

The calculations of the \ac{MLE} for $\hat{\theta}_{k}$, $\hat{\phi}_{j,k}$ and
$\hat{\psi}_{k}$ are now presented. The log likelihood of $\hat{\theta}$, $\hat{\phi}$ and $\hat{\psi}$ as derived
from the Equation~\eqref{eq:bhh:selection_mechanism:predictive_model_prop_to} is given in Equation~\eqref{eq:bhh:optimisation_step:mle:log_likelihood_all} below.

@@

Equation~\eqref{eq:bhh:optimisation_step:mle:log_likelihood_all} can be broken down into each of its factors. Consider the log likelihood of $\psi$  as denoted by Equation~\eqref{eq:bhh:optimisation_step:mle:log_likelihood_psi} below.

@@

The MLE for $\hat{\psi_{k}} $ can be calculated by taking the derivative of Equation~\eqref{eq:bhh:optimisation_step:mle:log_likelihood_psi} with respect to $\psi_{k}$ and equating to zero as presented in Equation~\eqref{eq:bhh:optimisation_step:mle:proof_log_likelihood_psi} as follows.

@@

The value for $\hat{\theta_{k}} $ can be calculated similarly. However, one has to compensate for the $K-1$ simplex $S$ by adding error factor $\epsilon$ to correct values for  $\theta$ where $\sum_{k}^{K} \theta_{k} \neq 1$. The new log likelihood function is then given in Equation~\eqref{eq:bhh:optimisation_step:mle:proof_log_likelihood_theta_part_1} below.

@@

Solving first for $\epsilon$ yields Equation~\eqref{eq:bhh:optimisation_step:mle:proof_log_likelihood_theta_part_2} below.

@@

Then solving for $\theta_{k}$ yields Equation~\eqref{eq:bhh:optimisation_step:mle:proof_log_likelihood_theta_part_3} as presented below.

@@

Substituting back into the Equation yields Equation~\eqref{eq:bhh:optimisation_step:mle:proof_log_likelihood_theta_part_4} below.

@@

The value for $\hat{\phi}_{j, k} $ is calculated very similarly. To compensate for the $K-1$ simplex $S$ an error factor $\lambda = (\lambda_{1}, \dots, \lambda_{J})$ is added. This follows in Equation~\eqref{eq:bhh:optimisation_step:mle:proof_log_likelihood_phi_part_1} as is presented below.

@@

Solving first for $\lambda_{j}$ yields Equation~\eqref{eq:bhh:optimisation_step:mle:proof_log_likelihood_phi_part_2} as is given below.

@@

Then solving for $\phi_{j,k}$ yields Equation~\eqref{eq:bhh:optimisation_step:mle:proof_log_likelihood_phi_part_3} below.

@@

The likelihood is presented in Equation~\eqref{eq:bhh:optimisation_step:map:bayesian_analysis_lamens} above. The event $H$ is a multinomial distribution with parameter $\theta$. It is known that the conjugate prior to a multinomial distribution is a Dirichlet distribution (see Chapter \ref{chap:probability}. The \textit{prior} probability distribution for  $\theta_{k}$ is thus presented in Equation~\eqref{eq:bhh:optimisation_step:map:map_theta_prior} below.

@@

Furthermore, the event $E$ is also a multinomial distribution with parameter $\phi$. The \textit{prior} probability distribution for $\phi_{j,k}$ is thus presented in Equation~\eqref{eq:bhh:optimisation_step:map:map_phi_prior} below.

@@

Finally, the event $C$ is a binomial distribution with parameter $\psi$. It is known that the conjugate prior to a binomial distribution is a Beta distribution (see Chapter \ref{chap:probability}). The prior probability distribution for $\psi_{k}$ is thus presented in Equation~\eqref{eq:bhh:optimisation_step:map:map_psi_prior} below.

@@

Putting the likelihood and priors together yields the posterior distribution as given in Equation~\eqref{eq:bhh:optimisation_step:map:map_psi_prior} below.

@@

From Equation~\eqref{eq:bhh:optimisation_step:map:posterior} above, it can be seen that the posterior distribution has the same form $\mathcal{A}(v)$ as the prior distribution. The concentration update operations can then be given in equations \ref{eq:bhh:optimisation_step:map:alpha_update_operation}, \ref{eq:bhh:optimisation_step:map:beta_update_operation}, \ref{eq:bhh:optimisation_step:map:gamma1_update_operation} and \ref{eq:bhh:optimisation_step:map:gamma2_update_operation} below.
